### 2025-06-13 manuscripts/10-JERP-v3/manuscript_main.tex
edit manuscript

A large body of literature provides evidence of substantial earnings gaps across race and ethnicity in the US \autocite{bayer2018divergent, charles2008prejudice}. Hispanic individuals constitute a large and growing portion of the US population. As this population grows, it is crucial to determine whether ethnic discrimination affects their employment prospects, wages, and career advancement. Understanding these labor market outcomes is critical, as they directly affect broader societal issues like assimilation and economic mobility, \autocite{chettyUnitedStatesStill2014, chettyEffectsExposureBetter2016,chettyFadingAmericanDream2017}. These factors serve as key indicators of how successfully Hispanic immigrants can navigate society and ascend the socioeconomic ladder.

In this paper, I answer the following questions. Does having a Hispanic last name affect educational outcomes? Does having a Hispanic last name affect labor market outcomes? I aim to show that comparing Hispanic White individuals to non-Hispanic White individuals might create an artificially higher earnings gap since these groups differ in many observable characteristics.\footnote{In this study, I define Hispanic individuals specifically as US-born persons who self-identify as White and have at least one parent born in a Spanish-speaking country.} \footnote{Observable characteristics refer to factors that can be measured and quantified, such as education level, work experience, and immigration status.} My analysis focuses on US-born children with one foreign-born parent (i.e., inter-ethnic).\footnote{This study's focus on US-born children with foreign-born parents means the findings may not generalize to Hispanic children with US-born parents, who likely face different socioeconomic and cultural circumstances. The analysis does not account for heterogeneity in immigrant characteristics across different Spanish-speaking countries of origin, such as variations in educational attainment, socioeconomic status, and gender-specific migration patterns, which could influence both parental selection into migration and subsequent child outcomes.} Others have attempted to compare how native-born White Hispanics fare compared to non-Hispanic Whites and foreign-born Hispanics. \cites{antman2020ethnic} have compared the health and educational outcomes of Hispanic Whites to non-Hispanic Whites and native-born Hispanics to foreign-born Hispanics. They find gaps in education and health between Hispanics and non-Hispanic Whites. They also find that native-born Hispanics are more likely than their foreign-born counterparts to report poor health.\footnote{For additional studies examining ethnic identity and outcomes among Hispanic populations, see \textcite{antman2020ethnic,antmanEthnicAttritionObserved2016,antmanEthnicAttritionObserved2016a,antmanEthnicAttritionAssimilation2020}, which collectively document the relationships between ethnic identification, assimilation patterns, and various socioeconomic outcomes.} \textcite{davilaChangesRelativeEarnings2008} found gaps in labor market outcomes between Hispanics and non-Hispanic Whites, which they attribute largely to differences in education, experience, immigration status, and regional differences. This paper builds on those studies by emphasizing how cultural assimilation and generational status further shape educational and labor market disparities for Hispanics.

Understanding discrimination against Hispanics in labor markets has far-reaching implications that extend beyond individual earnings. Labor market discrimination can create persistent intergenerational disadvantages by limiting economic mobility, reducing access to quality healthcare and education, and constraining residential choices; \autocite{chettyUnitedStatesStill2014, chettyEffectsExposureBetter2016,chettyFadingAmericanDream2017,bowles2002inheritance, djajic2003assimilation}. Discrimination may foster occupational sorting and segregation, as shown by recent task-based models of racial wage gaps that identify race-specific barriers in “Contact” tasks and explain the stagnation of these gaps post-1980 \autocite{hurst2024task}. These barriers may trap Hispanic families in cycles of lower socioeconomic status, as reduced earnings and employment opportunities limit their ability to invest in their children's human capital or build wealth through homeownership and savings.\footnote{See \textcite{bowles2002inheritance} for an analysis of the intergenerational transmission of economic status.}

Although previous research has identified earnings gaps between Hispanic and non-Hispanic Whites, this paper makes several important contributions. First, it develops a novel empirical strategy that better isolates the causal effect of Hispanic ethnicity on educational and labor market outcomes by comparing children from inter-ethnic families. This approach helps control for typically unobservable family background characteristics that might confound traditional analyses. Second, it provides new evidence on the specific role of Hispanic surnames in driving discrimination, offering insights into how ethnic signals influence educational attainment and labor market outcomes. Finally, by examining both education and employment outcomes, this study helps illuminate the channels through which ethnic discrimination may perpetuate economic disparities across generations and extends earlier findings by demonstrating how naming cues can compound structural inequalities and reinforce existing barriers.

This study builds and improves upon the traditional Oaxaca-Blinder-Kitagawa decomposition approach often used in labor market discrimination studies \autocite{kitagawa1955components, oaxaca1973male,blinder1973wage}. While these decomposition methods have provided valuable insights into earnings gaps between groups, including Hispanics and non-Hispanic Whites \autocite{davilaChangesRelativeEarnings2008}, they are limited in their ability to account for unobserved differences between the groups. My approach comparing children of inter-ethnic marriages, provides a more closely matched comparison group, allowing me to better isolate the effect of likely having a Hispanic last name on labor market outcomes.

The methodological improvement in this paper is particularly timely and relevant given the rapidly changing demographics of the United States. The US population is growing increasingly diverse, with significant implications for labor market dynamics and potential discrimination. The proportion of non-Whites has increased by more than 10 percentage points from 13 percent in 1995 to 23 percent in 2019.\footnote{The proportion of non-Whites and Hispanics is based on the author's calculations using the Current Population Survey (CPS).} Native-born White Hispanic men earn 21\% less than non-Hispanic White men, although there is evidence that a substantial portion of this gap is due to differences in education \autocite{duncan2006hispanics, duncan2018identifying, duncan2018socioeconomic}. These discrepancies may also stem from discrimination against Hispanics, which can lead to reduced job opportunities, lower wages, and hindered assimilation \autocite{chettyWhereLandOpportunity2014,bowles2002inheritance, djajic2003assimilation}.\footnote{This study focuses on US-born children of inter-ethnic unions, many of whom have at least one foreign-born parent. Assimilation concerns remain relevant because foreign-born parents may pass on cultural norms and language preferences, which can still shape these children's educational and labor market outcomes.}

 Using this novel approach, I find significant effects of Hispanic surnames on both educational and labor market outcomes. Individuals who likely have Hispanic last names face substantial disadvantages in educational attainment and earnings compared to their counterparts with likely non-Hispanic names. Specifically, individuals with presumably Hispanic last names complete 0.2 fewer years of education than those with non-Hispanic names, even when controlling for family background. Individuals with presumably Hispanic last names are also 1 percentage point more likely to be unemployed and earn 5 percentage points less than those with likely non-Hispanic names. These findings suggest that ethnic discrimination continues to play a role in shaping economic opportunities for Hispanics in the United States, highlighting the need for targeted policies to address these persistent disparities.

Factors that affect labor market outcomes, like skills and stereotypes, are unobservable to economists, so it is challenging to identify discrimination. One strategy used by researchers is audit or resume studies. \textcite{bertrand2004emily} conducted an audit study by sending employers identical resumes that differed only in the ethnic and racial signal of the applicant’s name (Black-sounding versus White-sounding). They found that resumes with Black-sounding names received substantially fewer callbacks than their White counterparts. Audit studies only observe callbacks, not wages. The  current study uses a method developed by \textcite{rubinstein2014pride}. The authors compared the children of mixed marriages between Sephardic and Ashkenazi people in Israel.\footnote{Ashkenazi and Sephardic are two distinct Jewish ethnic groups.} They found that workers with Sephardic last names earn substantially less than those with Ashkenazi last names. I compare children of Hispanic fathers and US born mothers (henceforth HW) and likely have a Hispanic-sounding last name to children of US born fathers and Hispanic mothers (WH). This approach accounts for the fact that  couples are likely to have similar income, schooling, and socioeconomic background, so these factors are largely controlled for \autocite{averettBetterWorseRelationship2008, averettEconomicRealityBeauty1996}.\footnote{For more on assortative mating see \autocite{beckerTheoryMarriagePart1973, beckerTheoryMarriagePart1974, beckerTreatiseFamily1993, browningCollectiveUnitaryModels2006, chiapporiFatterAttractionAnthropometric2012}} Children of HW and WH marriages have more similar observable characteristics than children of endogamous/homogamous marriages (i.e., US born fathers-US born mothers and Hispanic fathers-Hispanic mothers). In the US, children  from households with a Hispanic father and White mother  are overwhelmingly likely to  have their father’s Hispanic last name, allowing us to investigate how this ethnic signal affects annual log earnings.

The choice to include separate results for men and women is motivated by evidence that discrimination based on sex/gender and ethnicity operate through distinct mechanisms. As \textcite{bertrand2020gender} discusses, women's labor market outcomes are shaped by persistent gender norms and stereotypes, particularly around motherhood and caregiving. \textcite{antecol2002relative} found that young Mexican women earned 9.5\% less than young white women in the US in 1994, and like the current study, differences in education appeared to be the most likely explanation for this gap. For Black women, differences in labor force attachment appear to be the key driver of wage disparities. The way discrimination manifests can also evolve differently by gender over time: \textcite{goldin2004making} found that women's professional identity and career continuity are increasingly important factors in labor market outcomes, while \textcite{darity2015tour} argue that a stratification economics approach helps reveal how group-based hierarchies are maintained through both discriminatory practices and intergenerational resource transfers. Disaggregating the results by gender provides important insights into how the intersection of ethnic and gender discrimination shapes educational and economic opportunities.

The main identifying assumption of my empirical strategy depends on the assumption that people born to HW parents are similar to their WH peer in all observable---and unobservable---aspects and characteristics relevant to the labor market, so surname is the only difference between these groups. Children from US-born-Hispanic homes might benefit from more favorable family conditions than those from Hispanic-US-born homes, considering Whites statistically have higher socioeconomic status than Hispanics.  These factors introduce doubts regarding whether children from Hispanic-US-born families receive comparable familial support and influences, genetically or environmentally, as those from US-born-Hispanic families.

Previous studies have also used names as a proxy for race and ethnicity \autocite{fryer2004causes, bertrand2004emily}. \textcite{fryer2004causes} establish that names can be a predictor of a person's race. They found that having a Black-sounding name, after controlling for the home environment at birth, does not affect labor market outcomes. These early innovations in leveraging name-based identification highlight the importance of understanding how such signals can capture both conscious and unconscious bias, placing the current study within a broader framework of research on race, names, and economic outcomes.

Audit studies in education economics investigated the effect of racial and ethnic signals on access to education. \textcite{bergman2018education} found that students with Hispanic-sounding names received 2 percentage points fewer responses from schools than students with White sounding names. \textcite{janssen2022guidance} found that guidance counselors restricted Asian students from advanced opportunities. \textcite{gaddis2024racial} found significant discrimination in interactions between school principals and Hispanic and Chinese American families. Finally, \textcite{bourabain2023school} found more evidence of discrimination in access to education against students with underprivileged backgrounds in the Flemish education system. Discrimination can affect educational access and success through multiple channels: restricted access to advanced coursework, biased academic counseling, limited encouragement to pursue higher education, and subtle institutional barriers that may cause Hispanic students to feel unwelcome or unsupported in educational settings. While Hispanic college enrollment has increased substantially in recent decades, discrimination within educational institutions may still impede degree completion. The current paper contributes to this literature by providing evidence that discrimination in access to education could lead to lower earnings for Hispanics, and how parental background and marriage market selection can shape these outcomes in ways not previously explored by earlier studies.

The rest of the paper is organized as follows. In Section \ref{sec:data}, I describe the data used in this paper. In Section \ref{sec:emp_model}, I present the empirical strategy. In Section \ref{sec:results}, I present the results from the estimation of the two specifications. Finally, in Section \ref{sec:con1}, I conclude.

@@

I use the 1994–2019 CPS data set to study the effect of Hispanic surnames on labor market outcomes, because this time period contains data on parents’ place of birth, ethnicity, and race. The CPS does not provide data on parents’ characteristics, essential to determine the family background, but the 1960–2000 US Census data includes parents’ place of birth, race, and ethnicity. I employ this information to construct 'synthetic parents' using a method developed by \textcite{rubinstein2014pride} linking husbands and wives  in the census data. Assuming that parents have children 25–40 years old, I link these “parents” using parents’ places of birth, birth year of "children," and the parents’ places of birth.

For my main analysis, I construct different educational variables: years of education, and whether or not the person received a high school diploma, associate degree, and/or bachelor’s degree. For labor market outcomes, I construct an unemployment rate variable using the main CPS, log annual earnings using the ASEC, and log weekly earnings using the outgoing rotation. The unemployment indicator is a binary variable that equals 1 if someone is unemployed and 0 if employed, based on civilian labor force status. Log annual earnings is the log of total personal income from the CPS’ ASEC supplement. Log weekly earnings is the natural logarithm of weekly earnings for hourly workers from the outgoing rotation CPS supplement. I use the following controls: age, hours worked, and state-year fixed effects. I also include the parents’ characteristics from the synthetic parents data as controls in alternative specifications. These controls include income and education of the synthetic parents.

\subsection{Children of the four parental types}

I use the CPS for my primary analysis of the effect of a Hispanic surname on earnings. I restrict my sample to US citizens  aged 25–40 born between 1960 and 2000 in the US. Taking advantage of data on parents’ place of birth, I divide the sample into four groups depending on their parents’ ethnicity. Mothers or fathers are Hispanic if they were born in a Spanish-speaking country or Puerto Rico, and non-Hispanic White if they were born in the United States.\footnote{The list of Spanish-speaking countries includes: Argentina, Bolivia, Chile, Colombia, Costa Rica, Cuba, Dominican Republic, Ecuador, El Salvador, Guatemala, Equatorial Guinea, Honduras, Mexico, Nicaragua, Panama, Paraguay, Peru, Spain, Uruguay, Venezuela.} Therefore, an observation can be the product of four types of parents:
\begin{enumerate}
\item US-born father and US-born mother (hereafter WW) 
\item US-born father and Hispanic mother (hereafter WH)---the comparison group and the group that is likely to have a non-Hispanic-sounding last name
\item Hispanic father and US-born mother (hereafter HW)---the focus of this study and the group most likely to have a Hispanic-sounding last name
\item Hispanic father and Hispanic mother (hereafter HH).
\end{enumerate}

My sample includes inter-ethnic children with Hispanic ancestry aged 25–40 who are US citizens. I include both Hispanic White and non-Hispanic White individuals in this study. For second-generation immigrants, I use parents' place of birth as a proxy for ethnicity. This approach helps avoid biased estimates that could result from ethnic attrition, as demonstrated by \textcite{hadah2024effect}. \footnote{A person self-reports Hispanic identity by answering the Census question: ``Is this person Spanish/Hispanic/Latino?''} \footnote{Ethnic attrition happens when a US-born descendant of a Hispanic immigrant fails to self-identify as Hispanic. For more discussion of this phenomenon see \textcite{antmanEthnicAttritionObserved2016,antmanEthnicAttritionAssimilation2020}} To maintain analytical clarity, I restricted the sample to Hispanic and non-Hispanic White individuals, excluding other racial groups who also identify as Hispanic, to avoid confounding racial factors.

Regarding the classification of US-born parents as 'White'. While the United States has a long history of immigration, the probability that a US-born parent in this sample is a second-generation or later immigrant from a Spanish-speaking country is very low, as shown in Table \ref{tab:mat3}. The majority of Hispanics in the US during this period were first- and second-generation immigrants \autocite{antmanEthnicAttritionAssimilation2020}. Only 3\% of native-born Americans identified as Hispanic during this period, making it statistically unlikely that an interethnic child with a native-born parent is also a second-generation Hispanic immigrant. While this approach limits the generalizability of the findings to the broader Hispanic population, it provides a cleaner identification of surname effects, which is the primary focus of this study.

The CPS data distribution of the four types of parenthood is presented in Table \ref{tab:mat1}. The vast majority (96\%) are WW children. The second biggest group is HH, which constitutes 3\% of the sample. Inter-ethnic children, WH and HW, make up 1.35\% of the sample with 90,325 observations. Even though WH and HW are only 1.35\% of the sample, I have plenty of observations to carry out an analysis. The summary statistics for the children of the four types of marriages are presented in Table \ref{tab:c&p2}. Children of WW marriages (Column 1) do better on every measure while children of HH parents (Column 4) do worse than other children on every measure. Children of WH (Column 2) and HW (Column 3) marriages fall in between WW and HH children. The rates of self-reported Hispanic identity vary significantly across groups. Among children of WW marriages (Column 1), only 4\% of men and 5\% of women identify as Hispanic. This proportion increases substantially for children of inter-ethnic marriages: in WH families (Column 2), 74\% of men and 78\% of women identify as Hispanic, while in HW families (Column 3), the rates are 83\% for men and 81\% for women. Children of HH marriages (Column 4) show the highest rates of Hispanic identification, with 96\% of men and 97\% of women self-reporting as Hispanic.\footnote{The ethnic attrition rates are similar to those found in \textcite{antmanEthnicAttritionObserved2016,antmanEthnicAttritionAssimilation2020, hadah2024effect}.}

\subsection{Synthetic parents}

Using the 1960 to 2000 censuses, I constructed a data set of synthetic parents. The sample includes married White men and women. Even though the census asks a person whether they are Hispanic or not, I took advantage of the questions on place of birth to create a proxy for ethnicity. I consider a Hispanic persons who self-report White as their racial group and are born in a Spanish-speaking country. Consequently, US born in the sample are those who self-report White as race and are native-born. Using the information provided in the census, I can link husbands and wives with each other. I assume that parents have children between the ages of 25 and 40, so my sample consists of married White men and women with children born in the 1920 to 1975 cohorts\footnote{The construction of ``synthetic parents'' follows the method used by \textcite{rubinstein2014pride}.}.

@@

This aggregation process introduces some heterogeneity, as individual parents vary considerably from these group-level measures. However, this approach provides valuable information about typical family background characteristics that would otherwise be completely unobserved. The synthetic parents' characteristics serve as useful proxies for the socioeconomic environment in which these children were raised, allowing me to control for important background factors that influence educational and labor market outcomes. While more refined matching would be ideal, the publicly available data limit the potential matching dimensions. Nevertheless, the current approach provides meaningful estimates while acknowledging these data constraints.

I show the distribution of the four types of couples in Table \ref{tab:mat2}. US born husbands and US born wives (WW) make up the majority of couples in the sample, 96\% (5,141,737 couples). Hispanic husbands and wives (HH) are the second-largest group, compromising 2\% (119,749 couples). US born husbands and Hispanic wives (WH) $< 1\%$ (33,097 couples) of the sample and Hispanic husbands and US born wives (HW) are around 1\% (37,847 couples). I present the summary statistics of the parents in Table \ref{tab:synth}.

@@

Let $Y_{ist}$ be the outcome of interest for person $i$ in state $s$ at time $t$. $HW_{ist}$ is an indicator variable equal to 1 if person $i$ has a Hispanic father and US-born mother, and 0 if person $i$ has a US-born father and Hispanic mother. $X_{ist}$ is a vector of controls that includes age and hours worked, $\gamma_{st}$ represents state-year fixed effects, and $\phi_{ist}$ is the error term. The equation for this strategy is written as follows, and the sample is restricted to individuals from WH and HW families:

@@

$\beta_{1}$ is the coefficient of interest in this specification. $\beta_{1}$ represents the gap in outcomes between children of inter-ethnic marriages who likely have a Hispanic-sounding last name versus those who likely have a White-sounding last name. If $\beta_{1} > 0$, then people who likely have a Hispanic last name have better outcomes than people who likely have a White last name. If $\beta_{1} < 0$, then people who likely have a Hispanic last name have worse outcomes than people who likely have a White last name.

By comparing children from interethnic marriages, I can control for many unobservable family characteristics that may affect labor market outcomes, since these families are likely to be more similar to each other than to families with two Hispanic or two non-Hispanic White parents. This approach allows me to isolate the effect of surname, providing a more accurate estimate of its impact. The difference in means between Hispanic and non-Hispanic Whites could result from discrimination, but it could also be due to differences in innate abilities, skills, and parental investments . WH children will likely have a non-Hispanic last name, while HW children will likely have a Hispanic last name. This method was developed by \textcite{rubinstein2014pride}. This approach improves upon traditional Oaxaca-Blinder-Kitagawa decomposition methods, which are limited in their ability to account for unobserved differences between groups, by providing a more suitable comparison group to measure discrimination \autocite{oaxaca1973male,blinder1973wage,kitagawa1955components}.

The central assumption underpinning my estimation strategy is based on the hypothesis that individuals born to a HW person exhibit comparable characteristics to their peers of WH descent, especially in areas like educational background, skill sets, and work experiences that are significant determinants in employment opportunities, salary levels, and career advancement.

Two primary reasons underscore this assumption: 
\begin{itemize}
    \item First, there is significant selection in the marriage market. Since belonging to the Hispanic out-group may bring negative societal effects, it is worth exploring factors common among White women who have children with Hispanic men. 
    \item Second, fathers and mothers influence human capital accumulation differently \autocite{kimball2009risk,magruder2010intergenerational}. If marriage were random, then WH households might be a better environment for children. Using the CPS and the US Census, I will evaluate these empirical concerns.
\end{itemize}

A large body of empirical and theoretical work indicates that marriages exhibit strong selection, or `assortative matching' \autocite{averettBetterWorseRelationship2008, averettEconomicRealityBeauty1996, beckerTheoryMarriagePart1973, beckerTheoryMarriagePart1974, beckerTreatiseFamily1993, browningCollectiveUnitaryModels2006, chiapporiFatterAttractionAnthropometric2012}. Also, \textcite{duncanIntermarriageIntergenerationalTransmission2011} show a similar pattern of assortative matching among Mexicans in the United States. Interethnic WH and HW marriages generally consist of partners sharing similar characteristics, suggesting both WH and HW parents are similar in aspects relevant to labor market outcomes.

>>>>

Hispanic Americans represent one of the fastest growing demographic groups in the United States, making their labor market experiences increasingly central to understanding the American economy. Although extensive research documents persistent earnings gaps between racial and ethnic groups \autocite{bayer2018divergent, charles2008prejudice}, with native-born Hispanic White men earning 21\% less than non-Hispanic White men \autocite{duncan2006hispanics, duncan2018identifying, duncan2018socioeconomic}, the specific mechanisms affecting Hispanic workers' employment prospects, wages, and career advancement require closer examination. These labor market outcomes extend far beyond individual earnings—they fundamentally shape patterns of assimilation, economic mobility, and intergenerational progress \autocite{chettyUnitedStatesStill2014, chettyEffectsExposureBetter2016,chettyFadingAmericanDream2017}. Understanding how discrimination affects Hispanics offers valuable insight into both the obstacles they face and the possible routes for their socioeconomic advancement.

This paper examines whether having a Hispanic last name affects the educational and labor market outcomes of native-born Hispanics. I introduce a novel empirical strategy comparing children from interethnic families—those with likely Hispanic-sounding last names (HW) with those with White-sounding last names (WH). Specifically, I focus on US-born children with one foreign-born parent to isolate the effect of surname-based discrimination from other confounding factors.\footnote{Following the approach of \textcite{antman2020ethnic, hadah2024effect}, I define Hispanic individuals as people born in the US who self-identify as White and have at least one parent born in a Spanish-speaking country. This definition helps mitigate issues related to ethnic attrition. As demonstrated by \textcite{hadah2024effect}, ethnic attrition, particularly when associated with state-level bias, can result in an overestimation of the Hispanic-White gap in the most biased states. Additionally, focusing on White Hispanics provides cleaner estimates of discrimination based on Hispanic ethnicity by avoiding the confounding effects of racial discrimination that would complicate the interpretation of results.} This comparison leverages the similarity of family backgrounds in interethnic marriages to identify discrimination's role in shaping economic outcomes. Simple comparisons between Hispanic and non-Hispanic groups overstate earnings gaps, because these populations differ systematically in education, work experience, immigration status, and other observable characteristics.\footnote{Observable characteristics refer to factors that can be measured and quantified, such as education level, work experience, and immigration status.} By examining individuals who share similar backgrounds but differ primarily in their surnames, I can better isolate the impact of ethnic signals on educational and labor market outcomes.\footnote{This study's focus on US-born children with foreign-born parents means the findings may not generalize to Hispanic children with US-born parents, who likely face different socioeconomic and cultural circumstances. The analysis does not account for heterogeneity in immigrant characteristics in different Spanish-speaking countries of origin, such as variations in educational attainment, socioeconomic status, and gender-specific migration patterns, which could influence both parental selection in migration and subsequent child outcomes.}

This approach builds on \textcite{rubinstein2014pride}, who compared children of mixed Ashkenazi-Sephardic marriages in Israel, finding substantial wage penalties associated with Sephardic surnames. Previous work by \textcite{antman2020ethnic} and \textcite{davilaChangesRelativeEarnings2008} has shown that differences in education, experience, and immigration status explain much but not all of the Hispanic-White gap, suggesting discrimination may play a role. \textcite{antman2020ethnic} documents significant ethnic attrition among Hispanics and finds persistent gaps in education and health between native-born Hispanics and non-Hispanic Whites, with native-born Hispanics more likely to report poor health than their foreign-born counterparts.\footnote{For additional studies examining ethnic identity and outcomes among Hispanic populations, see \textcite{antman2020ethnic, antmanEthnicAttritionObserved2016, antmanEthnicAttritionObserved2016a, antmanEthnicAttritionAssimilation2020}, which collectively document the relationships between ethnic identification, assimilation patterns, and various socioeconomic outcomes.}

Discrimination against Hispanics in the US labor market undermines economic mobility and opportunity, reducing earnings and narrowing pathways to educational and professional advancement. These effects extend beyond wages to affect access to quality health care, residential choice, and occupational sorting \autocite{chettyUnitedStatesStill2014, hurst2024task}, potentially perpetuating inequality across generations. This paper's methodological improvement is particularly timely given the rapidly changing demographics of the United States. The proportion of non-Whites has increased by more than 10 percentage points from 13 percent in 1995 to 23 percent in 2019,\footnote{The proportion of non-Whites and Hispanics is based on the author's calculations using the Current Population Survey (CPS).} making it crucial to understand how ethnic signals affect labor market outcomes in an increasingly diverse society.

My empirical strategy relies on one key identifying assumption: children of Hispanic-father/White-mother couples (HW) and children of White-father/Hispanic-mother couples (WH) are alike on every trait that matters in the labor market, both observable and unobservable. If this holds, surname is the only systematic difference between them. This approach accounts for the fact that couples are likely to have similar income, schooling, and socioeconomic background, so these factors are largely controlled for \autocite{averettBetterWorseRelationship2008, averettEconomicRealityBeauty1996}.\footnote{For more on assortative mating see \textcite{beckerTheoryMarriagePart1973, beckerTheoryMarriagePart1974, beckerTreatiseFamily1993, browningCollectiveUnitaryModels2006, chiapporiFatterAttractionAnthropometric2012}} Indeed, children of HW and WH marriages have more similar observable characteristics than children of endogamous/homogamous marriages (that is, White-fathers/White-mothers and Hispanic fathers/Hispanic mothers), as I demonstrate in Section \ref{sec:hw-wh-couples-data}. It is probably true that offspring from interethnic unions have more comparable parental and cultural backgrounds with one another than those from endogamous Hispanic or White marriages do among themselves. In the US, children from households with a Hispanic father and White mother are overwhelmingly likely to have their father's Hispanic last name, allowing me to investigate how this ethnic signal affects labor market and educational outcomes \autocite{davenport2016role}.

My surname-based approach complements the audit study literature, which has documented discrimination but cannot observe actual wages. While \textcite{bertrand2004emily} found that Black-sounding names received fewer callbacks, and education audit studies show 2-11 percentage point penalties for Hispanic names in school responses \autocite{bergman2018education, gaddis2024racial}, my method reveals the downstream consequences of such discrimination on actual earnings and educational attainment. \textcite{fryer2004causes} established that names can be a predictor of a person's race, though they found that having a Black-sounding name, after controlling for the home environment at birth, does not affect labor market outcomes. These studies highlight the importance of understanding how such signals can capture both conscious and unconscious bias. The education audit experiments are particularly relevant, as discrimination can affect educational access through multiple channels: restricted access to advanced coursework as found by \textcite{janssen2022guidance} with Asian students, biased academic counseling, and subtle institutional barriers documented by \textcite{bourabain2023school} in the Flemish education system. While Hispanic college enrollment has increased substantially in recent decades, discrimination within educational institutions may still impede degree completion, ultimately affecting labor market outcomes.

As gender and ethnic discrimination operate through distinct mechanisms, I separate out a portion of my results based on gender. As \textcite{bertrand2020gender} notes, women's outcomes in the labor market are shaped by persistent norms around caregiving and motherhood. \textcite{antecol2002relative} found that in 1994, young Mexican women earned 9.5\% less than young White women in the US, with differences in education largely explaining the gap. \textcite{goldin2004making} emphasizes the growing importance of professional identity and career continuity for women, while \textcite{darity2015tour} show how group-based hierarchies are perpetuated through both discrimination and intergenerational transfers. The breakdown of my results by gender therefore reveals how ethnic and gender biases intersect to shape educational and economic opportunities.

The remainder of the paper is organized as follows. In Section \ref{sec:data}, I describe the data used in this paper. In Section \ref{sec:emp_model}, I present the empirical strategy. In Section \ref{sec:results}, I present the results of the estimation of the two specifications. Finally, in Section \ref{sec:con1}, I conclude.

@@

I use the 1994–2019 CPS data to study the effect of Hispanic surnames on labor market outcomes, as this period contains data on parents' place of birth. The place of birth of mothers and fathers is explicitly asked of all participants in the CPS starting 1994, and this information is essential for identifying interethnic marriages. Since the CPS lacks direct parental characteristics data, I construct 'synthetic parents' using the 1960–2000 US Census data, following \textcite{rubinstein2014pride}.\footnote{For more on 'synthetic' parents see \textcite{aaronson2008intergenerational}.} This method links married couples in census data based on parents' places of birth and children's birth years, assuming that parents have children between ages 25 and 40.

\subsection{Sample Construction}

My sample includes US citizens aged 25 to 40 years born between 1960 and 2000. Using parents' place of birth data, I classify observations into four parental types:

\begin{enumerate}
\item \textbf{WW}: US-born father and US-born mother
\item \textbf{WH}: US-born father and Hispanic mother (comparison group, likely non-Hispanic surname)
\item \textbf{HW}: Hispanic father and US-born mother (treatment group, likely Hispanic surname)
\item \textbf{HH}: Hispanic father and Hispanic mother
\end{enumerate}

Parents are classified as Hispanic if born in a Spanish-speaking country or Puerto Rico, and non-Hispanic White if US-born.\footnote{Spanish-speaking countries include: Argentina, Bolivia, Chile, Colombia, Costa Rica, Cuba, Dominican Republic, Ecuador, El Salvador, Guatemala, Equatorial Guinea, Honduras, Mexico, Nicaragua, Panama, Paraguay, Peru, Spain, Uruguay, Venezuela.} This approach helps avoid biased estimates from ethnic attrition \autocite{hadah2024effect}.\footnote{Ethnic attrition occurs when US-born descendants of Hispanic immigrants fail to self-identify as Hispanic. See \textcite{antmanEthnicAttritionObserved2016,antmanEthnicAttritionAssimilation2020}.}

@@

This aggregation process introduces some heterogeneity, as individual parents vary considerably from these group-level measures. However, this approach provides valuable information about typical family background characteristics that would otherwise be completely unobserved \autocite{rubinstein2014pride,aaronson2008intergenerational}. The synthetic parents' characteristics serve as useful proxies for the socioeconomic environment in which these children were raised, allowing me to control for important background factors that influence educational and labor market outcomes. While more refined matching would be ideal, the publicly available data limit the potential matching dimensions. Nevertheless, the current approach provides meaningful estimates while acknowledging these data constraints.

The sample is restricted to Hispanic and non-Hispanic White individuals to avoid confounding racial factors. While this limits generalizability to the broader Hispanic population, it provides cleaner identification of surname effects—the study's primary focus. While the United States has a long history of immigration, the probability that a US-born parent in this sample is a second-generation or later immigrant from a Spanish-speaking country is very low, as shown in Table \ref{tab:mat3}. The majority of Hispanics in the US during this period were first- and second-generation immigrants. Only 3\% of native-born Americans identified as Hispanic during this period, making it statistically unlikely that an interethnic child with a native-born parent is also a second-generation Hispanic immigrant.

\subsection{Variables and Measures}

\textbf{Educational outcomes}: Years of education, high school diploma, associate degree, and bachelor's degree completion.

\textbf{Labor market outcomes}:
\begin{itemize}
\item Unemployment (binary indicator from civilian labor force status)
\item Log annual earnings (from ASEC supplement)
\item Log weekly earnings (from outgoing rotation for hourly workers)
\end{itemize}
\textbf{Controls}: Age, hours worked, state-year fixed effects, and synthetic parents' characteristics (income and education) in alternative specifications.

\subsection{Sample Distribution}
Table \ref{tab:mat1} shows that WW children comprise 96\% of the sample, HH children 3\%, and interethnic children (WH and HW) make up 1.35\% with 90,325 observations—sufficient for robust analysis. Summary statistics and group comparisons are discussed in Section \ref{sec:hw-wh-couples-data}.

@@

Let $Y_{ist}$ be the outcome of interest for person $i$ in state $s$ at time $t$. $HW_{ist}$ is an indicator variable equal to one if person $i$ has a Hispanic father and US-born mother, and zero if person $i$ has a US-born father and Hispanic mother. $X_{ist}$ is a vector of controls that includes age, hours worked, and synthetic parents' characteristics (income and education), $\gamma_{st}$ represents state-year fixed effects, and $\phi_{ist}$ is the error term. The equation for this strategy is written as follows, and the sample is restricted to individuals from WH and HW families:

@@

$\beta_{1}$ is the coefficient of interest in this specification. $\beta_{1}$ represents the gap in outcomes between children of inter-ethnic marriages who likely have a Hispanic-sounding last name versus those who likely have a White-sounding last name. If $\beta_{1} > 0$, then people who likely have a Hispanic last name have better outcomes than people who likely have a White last name. If $\beta_{1} < 0$, then people who likely have a Hispanic last name have worse outcomes than people who likely have a White last name. This strategy follows \textcite{rubinstein2014pride}, comparing children from WH and HW interethnic families to estimate the effect of surname-based ethnic cues. As detailed in the Introduction, this isolates discrimination effects by holding family background constant.

The central assumption underpinning my estimation strategy is that individuals born to a HW person exhibit characteristics comparable to their peers of WH descent, especially in areas like educational background, skill sets, work experiences, family culture, and parenting styles, that are significant determinants in employment opportunities, salary levels, and career advancement.

A potential concern is measurement error in using CPS data to infer parental ethnicity, since it relies on place of birth rather than self-identified Hispanic origin. However, during the period studied, second-generation Hispanics were rare: only about 3\% of native-born Americans identified as Hispanic, see Table \ref{tab:mat3}. This makes it unlikely that US-born parents in interethnic unions are second-generation+ Hispanic immigrants.

Prior research documents strong assortative mating in interethnic unions, particularly across education and income lines \autocite{beckerTheoryMarriagePart1973, chiapporiFatterAttractionAnthropometric2012, duncanIntermarriageIntergenerationalTransmission2011}. I confirm this pattern using synthetic parent data in Section \ref{sec:hw-wh-couples-data}.

It is important to address potential statistical inference issues related to using synthetic parents' characteristics as control variables. The parental characteristics are included solely as controls and not for inference purposes. I do not interpret these coefficients, as my primary interest lies in $\beta_1$, which captures the gap between Hispanic and White last names. Since I am using these synthetic characteristics purely as controls, without drawing any conclusions from their coefficients, this should not affect the standard errors or p-values of my coefficient of interest $\beta_1$.

The synthetic parents are constructed using the birth year of the child and are not based on actual parents in the CPS sample at the time of the survey. The place of birth of parents—mothers and fathers—are questions asked of all participants in the Current Population Survey (CPS) starting 1994. Therefore, I do not use the information of parents who are in the CPS sample at the time of the survey, but rather construct synthetic parents from Census data. This approach means that concerns about non-random attrition of parents from the CPS sample do not apply to my analysis.
